\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-07 chapters/bci.tex
checking bttda

\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\todo{search for ??}
\todo{acro usage in chapters}

\chapter{Brain-computer interfaces}
\todo{are textcites used when necessary?}
\label{sec:bci}
People with \ac{sspi} due to
neurodegenenerative disease, stroke, traumatic brain or spine injury
become entirely reliant on caregivers or family members in their day to day
routine.
Wouldn't it be great if we could help them regain some quality of life by
making them more independent?

The most basic ability any independent human being needs, is interaction with
their environment.
In its core essence, this is the ability to communicate your intentions,
emotions, frustrations or thoughts such that they might be acted on.
For an able-bodied person, physical interaction happens through body movements
while communication is usually done through speech and body language
Both require sending signals through the body's nervous system to control
muscles.
Not everyone has this capability.

The key problem for severely paralyzed patients is that the mind wants to go
where the body cannot follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton.
The term \ac{bci} is usually preferred for
assistive technology and communication devices.
},
a system that directly couples actions in the `real world' to the mental state of
the user.

Starting in the 1970's from methods to establish basic communication from
minimal brain activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities
by bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab
setting and can be used as assistive communication technology by patients with
impaired speech~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting edge experimental examples include a brain-spine
interface allowing a paraplegic patient to walk again\cite{Lorach2023},
a \ac{bci} translating internal speech to a facial avatar speaking the imagined
text~\cite{Metzger2023}, and fast communication through decoding imagined
handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

In general, devices processing direct inputs from the central nervous system
are useful for rehabilitation and medical diagnosis or treatment.
Additionally, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality~\cite{SiMohammed2017}.
Ultimatly, they have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
It is this \ac{bci} controlled communication is where our focus lies,
as it offers a transformative means for individuals with severe motor
impairments to regain functional independence to write and speak through
assistive technology.
This allows them to talk to their loved ones and caregivers, exercise hobbies
and makes many aspects of daily life more accessible.

@@

Simply put, a \ac{bci} records the user's brain activity, extracts some
relevant output from this brain activity and couples this output to a function
of a device as illustrated in Figure~\ref{fig:bci/loop}
Optionally, the user can then observe the action of the device and adapt their
behavior or brain activity accordingly, closing the loop\footnote{%
  \textcite{BCISociety2024} has recently formalized this into the
  following definition:
  \it``A brain-computer interface is a system that measures brain activity and
  converts it in (nearly) real-time into functionally useful outputs to replace,
  restore, enhance, supplement, and/or improve the natural outputs of the brain,
  thereby changing the ongoing interactions between the brain and its external or
  internal environments. It may additionally modify brain activity using targeted
  delivery of stimuli to create functionally useful inputs to the
  brain.''
}.

@@

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what is going on in the brain.
Multiple signals can be recorded from the brain, including bloodflow,
magnetical electrical activity.
For several reasons laid out in~\ref{sec:bci-recording}, the electrical modality
is especially interesting.

The recorded brain activity forms only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
Instead, it would help greatly if we knew which specific activity to look for.
A \ac{bci} operates according to a specified \emph{paradigm}, which defines the
method by which interaction is conducted.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
to couple specific activity in the brain to specific information or actions we
want to convey via the \ac{bci}.
The brain singal now \emph{modulates} this information.
In turn, the detection of this signal can be coupled to an action, like
typing the inteded letter A or B in Figure~\ref{fig:bci/loop}.
The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci/paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific
brain signatures related to our paradigm need to be identified within the recorded
activity.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own brain activity.
On the other hand, the activity of interest often
originates from a specific region or network within the brain.
This activity might not easily be discerned from all other brain and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing} (see Section~\ref{sec:bci/preprocessing}) and by choosing
the correct representation of the signal (i.e. \emph{feature extraction}, but the problem
of retrieving or \emph{decoding} only the relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder and interpreted in function of the
desired \ac{bci} output.
Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.
For a communication system, this means coupling the decoded information to
e.g.\ a virtual keyboard or speech synthesizer.
On the other hand, the actions performed by the \ac{bci} can themselves alter
the user's cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like tactile
feedback for movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural
pathways, can have a positive impact on \ac{bci} performance and gives rise to
rehabilitation applications.

To summarize, a \ac{bci} can facilitate direct interaction central nervous system.
To do this, information is modulated in the brain signal according to the
\ac{bci} paradigm by stimulating, the user or having them perform a task.
Their brain signal is subsequently recorded and preprocessed to clean it.
Relevant features that contain aspects of the modulated signal are extracted,
and the modulate information is retrieved by the decoder, a machine learning
model.
This information can then acted on and the loop is closed if this action
affects the user.
Let us take a look at each of these elements separately in the next section.

@@

The technologies above all measure bloodflow (the hemodynamic signal), which only indirectly
reflects brain activity.
This signal reacts too slowly to reflect the real-time activity that is
of interest in \ac{bci} for communication, and it carries little information
other than the brain region from where it is originating.

A better candidate is the neuronal electrical activity.
The brain contains 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential}
is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.
The firing of a neuron, or the combined firing of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
The related discipline is \emph{electrophysiology}.

Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality of electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials and
postsynaptic potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary
equipment is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.
As a consequence, the \ac{bci} field relies heavily on the
electrical activity of the brain.

\subsection{Invasive electrophysiology}
The \emph{invasiveness} of the technology forms an
important concept in determining its suitability for a given application and
patient.
As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often
measure a very specific brain region and can result in better signal quality, at the cost of the risks
involved with brain surgery.

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  Various electrophysiology recording technologies.
}{%
	Various electrophysiology recording technologies and \\
	their respective position in a cross-section of the skull. \\
	There is a trade-off between invasiveness and  \\
	spatial resolution.
}{%
  fig:bci/recording
}

Specifically, invasive methods are less noisy and come with an increased \emph{spatial resolution},
meaning they can extract a signal from a specific brain region or even a set of
neurons of interest.
This makes it easier to retrieve a signal with known origin from the brain,
which might be leveraged by the paradigm to modulate information.
Generally, microelectrodes penetrating the cortex are considered to have the highest
spatial resolution.
These are needles with a diameter smaller than 10µm and one or multiple
electrodes that extend several millimmeters into the cortex.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular
potential of a small group of neurons, or even intracellular single neuron
action potential spikes.
They come in single-electrode form, or, more popularly, in microelectrode
arrays.
Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
NeuraLink implant~\cite{Musk2019} and IMEC's Neuropixels
2.0~\cite{Steinmetz2021}.

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Depth electrodes are longer rods with multiple electrodes than can extend
deeper into the brain.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

Some impressive recent breakthroughs in speech and motor \ac{bci} for communication
have been realized using
invasive mictroelectrode arrays~\cite{Willett2021} and $\mu$\Ac{ecog}\cite{Metzger2023}.
Furthermore, recent advances in recording technology focus on improving implant
durability and resolution~\cite{Steinmetz2021}, and on balancing the invasiveness
tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
through the bloodstream via a catheter, removing the need for open brain
surgery.
Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
potential communication \ac{bci} user groups consistently show that non-invasive
technology is preferred over implanted electrodes, unless the added value of invasive BCI is
sufficiently large~\cite{Kageyama2020}.
This provides justification for focusing part of the current research effort on
non-invasive BCI.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
neurology practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
offers great temporal resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volt (V), often on the scale of microvolt ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical
consumer-grade \ac{eeg} headsets exist in the form of (often wireless) electrode
caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and
research systems often use active electrodes with preamplifiers and
electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality, since they require a trained
individual to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

The major drawback of \ac{eeg} is its poor spatial resolution~\cite{Ferree2001}.
Since they sit on the outside of the scalp electrodes are large and distance
relative to the neurons in the cortex they ought to measure.
Furthermore, the layers of skull and cerebrospinal fluid in between can
attenuate the signal and cause volume conduction~\cite{Broek1998}, which results in electrodes
measuring a mixture of activity from sources elsewhere in the brain.
This also has a negative effect on spectral resolution.
The noise problem mentioned earlier is also prominently present in \ac{eeg}
recordings.
In addition to picking up brain activity other than that of the region of
interest, \ac{eeg} also records other artifacts, from the environment, the power
supply, nearby electrical equipment, or the user's muscle
movements~\cite{Urigueen2015}.
Combined, this means \ac{eeg} is inherently noisy.

@@

The \ac{bci} paradigm~\cite{Xu2021,Neeling2019} is the key to translating brain activity into useful
output actions, since it defines which neural responses will be elicited and
captured as features.
A paradigm can be loosely seen as \emph{a way of interacting with the
\ac{bci}}.
According to \textcite{Zander2011}, they can be arranged by their reliance on
external stimulation and the engagement of the user as shown in
Figure~\ref{fig:bci/paradigms}.

@@

Paradigms that require little active participation of the user like
redirecting attention or initiating imagined actions are classified as \emph{passive
\ac{bci}}.
From the user's point of view, this concept is probably the most attractive,
since their cognitive state state would be inferred from just
their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like
emotion or affect monitoring~\cite{Torres2020,Libert2019, Muehl2014}, workload
monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
detection~\cite{SiMohammed2020}.
It also has clinical applications in the form of neurofeedback~\cite{Hammond2011}.
Nevertheless, active participation facilitates training data collection for
supervised machine learning since the conditions can be more easily controlled
by the user or the stimulation.

Paradigms requiring high active participation can be split up in \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user,
such as imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive
communication methods often limit the considered domain to a few movement
directions to control a cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded
like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
They might require extensive user training and are sensitive to
correct task instruction.
Imagined speech or movements can be performed in a multitude of ways, which
can each have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing
poor performance and giving rise to the concept of \ac{bci}
illiteracy~\cite{Allison2010}\footnote{%
Recently, the concept of BCI illiteracy is increasingly being seen as outdated.
Instead of attributing poor performance to the user's inherent limitations,
critiques point out that the issue lies more with the design of the
systems themselves. BCI systems often fail to account for individual
variability in brain anatomy and cognitive strategies. As a result, what has
been labeled as `illiteracy' may, in fact, be a reflection of systems that are
not flexible enough to adapt to different users. This perspective
shifts the responsibility to the individual, rather than recognizing that
better, more adaptive systems could overcome many of these performance
challenges, thus standing in the way of progress.
Moreover, the term `illiteracy' itself carries negative
connotations, implying a user deficiency that reinforces normative assumptions,
when in reality, a more inclusive approach to design could enhance BCI
accessibility for a wider range of individuals. This evolving view calls for a
reframing of the problem, focusing on improving the adaptability of BCIs rather
than labeling users as illiterate~\cite{Becker2022,Thompson2019}.
}.

\subsection{Reactive \acsp{bci}}
Reactive \ac{bci} take another approach, by providing a discrete set of
sensory stimulations towards which the user's attention can directed.
Compared to active \ac{bci}, reactive paradigms can more easily induce fatigue
due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

>>>>

\todo{comments liesel paradigms}
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\todo{search for ??}
\todo{acro usage in chapters}
\todo{are textcites used when necessary?}

\chapter{Brain-computer interfaces}
\label{sec:bci}
People with \ac{sspi} due to neurodegenerative disease, stroke, traumatic brain or spine
injury become entirely reliant on caregivers or family members in their day-to-day
routine.
Wouldn't it be great if we could help them regain some quality of life by making them
more independent?

The most basic ability any independent human being needs is interaction with their
environment.
In its core essence, this is the ability to communicate intentions, emotions,
frustrations, or thoughts such that they might be acted on.
For an able-bodied person, physical interaction happens through body movements, while
communication is usually done through speech and body language.
Both require sending signals through the body's nervous system to control muscles.
Not everyone has this capability.

The key problem for severely paralyzed patients is that the mind wants to go where the
body cannot follow.
Couldn't we, therefore, design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of technology, like
a robot arm or speech synthesizer, interacting with the world on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also termed brain-machine interface
(BMI) when coupled to a physical actuator, like a robot arm or an exoskeleton.
The term \ac{bci} is usually preferred for assistive technology and communication
devices.}, a system that directly couples actions in the `real world' to the mental state
of the user.

Starting in the 1970s from methods to establish basic communication from minimal brain
activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities by
bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
be used as assistive communication technology by patients with impaired
speech~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting-edge experimental examples include a brain-spine interface allowing a
paraplegic patient to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
communication through decoding imagined handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

In general, devices processing direct inputs from the central nervous system are useful
for rehabilitation and medical diagnosis or treatment.
Additionally, they bring a new paradigm to human-computer interaction, especially when
paired with virtual or augmented reality~\cite{SiMohammed2017}.
Ultimately, they have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
It is this \ac{bci}-controlled communication where our focus lies, as it offers a
transformative means for individuals with severe motor impairments to regain functional
independence to write and speak through assistive technology.
This allows them to talk to their loved ones and caregivers, exercise hobbies, and makes
many aspects of daily life more accessible.

@@

Simply put, a \ac{bci} records the user's brain activity, extracts some relevant output
from this brain activity, and couples this output to a function of a device as
illustrated in \autoref{fig:bci/loop}.
Optionally, the user can then observe the action of the device and adapt their behavior
or brain activity accordingly, closing the loop\footnote{\textcite{BCISociety2024} has
recently formalized this into the following definition: \it``A brain-computer interface
is a system that measures brain activity and converts it in (nearly) real-time into
functionally useful outputs to replace, restore, enhance, supplement, and/or improve the
natural outputs of the brain, thereby changing the ongoing interactions between the
brain and its external or internal environments.
It may additionally modify brain activity using targeted delivery of stimuli to create
functionally useful inputs to the brain.''}.

@@

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of what is
going on in the brain.
Multiple signals can be recorded from the brain, including blood flow and magnetic or
electrical activity.
For several reasons laid out in \autoref{sec:bci-recording}, the electrical modality is
especially interesting.

The recorded brain activity forms only one part of the interaction scheme presented in
\autoref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain activity
to try to extract the desired output.
Instead, it would help greatly if we knew which specific activity to look for.
A \ac{bci} operates according to a specified \emph{paradigm}, which defines the method
by which interaction is conducted.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research to couple
specific activity in the brain to specific information or actions we want to convey via
the \ac{bci}.
The brain signal now \emph{modulates} this information.
In turn, the detection of this signal can be coupled to an action, like typing the
intended letter A or B in \autoref{fig:bci/loop}.
The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the exact
task performed by the user, and is often closely linked to the manner of feedback in the
case of a closed-loop system.
As shown in \autoref{sec:bci/paradigms}, there is a wide variety of \ac{bci} paradigms
based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific brain
signatures related to our paradigm need to be identified within the recorded activity.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks,' all of which generate their own brain activity.
On the other hand, the activity of interest often originates from a specific region or
network within the brain.
This activity might not easily be discerned from all other brain and environmental
activity.
Compare this to trying to listen to a single speaker at a noisy party where everyone is
shouting over each other.
Some obvious sources of interference can be easily filtered out by \emph{preprocessing}
(see \autoref{sec:bci/preprocessing}) and by choosing the correct representation of the
signal (i.e., \emph{feature extraction}), but the problem of retrieving or
\emph{decoding} only the relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm can be retrieved
from it by the decoder and interpreted in function of the desired \ac{bci} output.
\autoref{sec:bci/decoding} takes a closer look at some common decoding techniques.

Finally, the loop can be closed by coupling this to a device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact with
their environment.
For a communication system, this means coupling the decoded information to, e.g., a
virtual keyboard or speech synthesizer.
On the other hand, the actions performed by the \ac{bci} can themselves alter the user's
cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of Parkinson's
disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like tactile feedback for
movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions back to the
user.
The user's brain will then adapt through reinforcement learning, causing changes in
behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural pathways,
can have a positive impact on \ac{bci} performance and gives rise to rehabilitation
applications.

To summarize, a \ac{bci} can facilitate direct interaction with the central nervous
system.
To do this, information is modulated in the brain signal according to the \ac{bci}
paradigm by stimulating the user or having them perform a task.
Their brain signal is subsequently recorded and preprocessed to clean it.
Relevant features that contain aspects of the modulated signal are extracted, and the
modulated information is retrieved by the decoder, a machine learning model.
This information can then be acted on, and the loop is closed if this action affects the
user.
Let us take a look at each of these elements separately in the next section.

@@

The technologies above all measure blood flow (the hemodynamic signal), which only
indirectly reflects brain activity.
This signal reacts too slowly to reflect the real-time activity that is of interest in
\ac{bci} for communication, and it carries little information other than the brain
region from where it is originating.

A better candidate is the neuronal electrical activity.
The brain contains 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential} are electrical
pulses occurring when a neuron receives input from another neuron and is activated.
The firing of a neuron, or the combined firing of groups of synchronized neurons, thus
generates an electrical field in and near the brain, which can be measured using
electrodes.
The related discipline is \emph{electrophysiology}.

Because of the desirable high temporal resolution~\cite{Easttom2021} and practicality of
electrophysiology, we will focus only on methods to record electromagnetic activity
originating from neuronal action potentials and postsynaptic potentials.
The magnetic field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary equipment
is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of lab
conditions.
As a consequence, the \ac{bci} field relies heavily on the electrical activity of the
brain.

\subsection{Invasive electrophysiology}

The \emph{invasiveness} of the technology forms an important concept in determining its
suitability for a given application and patient.
As illustrated in \autoref{fig:bci/recording}, invasive electrodes can often measure a
very specific brain region and can result in better signal quality, at the cost of the
risks involved with brain surgery.

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  Various electrophysiology recording technologies.
}{%
	Various electrophysiology recording technologies and their respective position in a
	cross-section of the skull.
	There is a trade-off between invasiveness and spatial resolution.
}{%
  fig:bci/recording
}

Specifically, invasive methods are less noisy and come with an increased \emph{spatial
resolution}, meaning they can extract a signal from a specific brain region or even a
set of neurons of interest.
This makes it easier to retrieve a signal with a known origin from the brain, which
might be leveraged by the paradigm to modulate information.
Generally, microelectrodes penetrating the cortex are considered to have the highest
spatial resolution.
These are needles with a diameter smaller than 10µm and one or multiple electrodes that
extend several millimeters into the cortex.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular potential
of a small group of neurons, or even intracellular single neuron action potential
spikes.
They come in single-electrode form, or, more popularly, in microelectrode arrays.
Well-known examples are the Utah array~\cite{Maynard1997}, the more recent NeuraLink
implant~\cite{Musk2019}, and IMEC's Neuropixels 2.0~\cite{Steinmetz2021}.

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes such as those used in stereo-\ac{eeg} \ac{bci}~\cite{Wu2024} tasks and
closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
symptoms~\cite{Arlotti2018}.
Depth electrodes are longer rods with multiple electrodes that can extend deeper into
the brain.
Since these still penetrate the cortex, electrodes that sit on top of the cortex are
more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and
epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019}, with a higher
number of smaller electrodes per surface area, allow for more precise measurements.

Some impressive recent breakthroughs in speech and motor \ac{bci} for communication have
been realized using invasive microelectrode arrays~\cite{Willett2021} and
$\mu$\Ac{ecog}~\cite{Metzger2023}.
Furthermore, recent advances in recording technology focus on improving implant
durability and resolution~\cite{Steinmetz2021}, and on balancing the invasiveness
tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered through the
bloodstream via a catheter, removing the need for open brain surgery.
Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
potential communication \ac{bci} user groups consistently show that non-invasive
technology is preferred over implanted electrodes, unless the added value of invasive
BCI is sufficiently large~\cite{Kageyama2020}.
This provides justification for focusing part of the current research effort on
non-invasive BCI.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical neurology
practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and offers
great temporal resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volts (V), often on the scale of microvolts ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical consumer-grade
\ac{eeg} headsets exist in the form of (often wireless) electrode caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and research
systems often use active electrodes with preamplifiers and electrolyte gel to reduce
impedance for improved signal quality, which slightly decreases practicality, since they
require a trained individual to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both patients and
researchers.

The major drawback of \ac{eeg} is its poor spatial resolution~\cite{Ferree2001}.
Since they sit on the outside of the scalp, electrodes are large and distant relative to
the neurons in the cortex they ought to measure.
Furthermore, the layers of skull and cerebrospinal fluid in between can attenuate the
signal and cause volume conduction~\cite{Broek1998}, which results in electrodes
measuring a mixture of activity from sources elsewhere in the brain.
This also has a negative effect on spectral resolution.
The noise problem mentioned earlier is also prominently present in \ac{eeg} recordings.
In addition to picking up brain activity other than that of the region of interest,
\ac{eeg} also records other artifacts from the environment, the power supply, nearby
electrical equipment, or the user's muscle movements~\cite{Urigueen2015}.
Combined, this means \ac{eeg} is inherently noisy.

@@

The \ac{bci} paradigm~\cite{Xu2021,Neeling2019} is the key to translating brain activity
into useful output actions, since it defines which neural responses will be elicited and
captured as features.
A paradigm can be loosely seen as \emph{a way of interacting with the \ac{bci}}.
According to \textcite{Zander2011}, they can be arranged by their reliance on external
stimulation and the engagement of the user as shown in \autoref{fig:bci/paradigms}.

@@

Paradigms that require little active participation from the user, like redirecting
attention or initiating imagined actions, are classified as \emph{passive \ac{bci}}.
From the user's point of view, this concept is probably the most attractive, since their
cognitive state would be inferred from just their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like emotion or affect
monitoring~\cite{Torres2020,Libert2019, Muehl2014}, workload monitoring~\cite{Zanetti2021},
or, more dependent on stimulation, error detection~\cite{SiMohammed2020}.
It also has clinical applications in the form of neurofeedback~\cite{Hammond2011}.
Nevertheless, active participation facilitates training data collection for supervised
machine learning, since the conditions can be more easily controlled by the user or the
stimulation.

Paradigms requiring high active participation can be split up into \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user, such as
imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive communication
methods often limit the considered domain to a few movement directions to control a
cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded activities like
natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
They might require extensive user training and are sensitive to correct task
instruction.
Imagined speech or movements can be performed in a multitude of ways, which can each
have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing poor
performance and giving rise to the concept of \ac{bci} illiteracy~\cite{Allison2010}\footnote{%
Recently, the concept of BCI illiteracy is increasingly being seen as outdated.
Instead of attributing poor performance to the user's inherent limitations, critiques
point out that the issue lies more with the design of the systems themselves.
BCI systems often fail to account for individual variability in brain anatomy and
cognitive strategies.
As a result, what has been labeled as `illiteracy' may, in fact, be a reflection of
systems that are not flexible enough to adapt to different users.
This perspective shifts the responsibility to the individual, rather than recognizing
that better, more adaptive systems could overcome many of these performance challenges,
thus standing in the way of progress.
Moreover, the term `illiteracy' itself carries negative connotations, implying a user
deficiency that reinforces normative assumptions, when in reality, a more inclusive
approach to design could enhance BCI accessibility for a wider range of
individuals.
This evolving view calls for a reframing of the problem, focusing on improving the
adaptability of BCIs rather than labeling users as illiterate~\cite{Becker2022,Thompson2019}.}.

\subsection{Reactive \acsp{bci}}
Reactive \ac{bci} take another approach by providing a discrete set of sensory
stimulations towards which the user's attention can be directed.
Compared to active \ac{bci}, reactive paradigms can more easily induce fatigue due to
the constant stimulation.
These paradigms are also somewhat less intuitive compared to, e.g., speech, and their
expressive power is limited by how many different stimuli can effectively be presented
and attended to within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.
